---
title: "Final Project - NBA Exploratory Data Analysis"
author: "Clajerson Gimena"
date: "6-8-2025"
output:
  pdf_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  comment = "", 
  message = FALSE, 
  warning = FALSE, 
  fig.align = "center"
)
```

### Load Packages

```{r, message = FALSE}
library(RSQLite)
library(DBI)
library(RMariaDB)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
library(bit64)
library(tidyr)
library(car)
library(gridExtra)
library(grid)
library(patchwork)
```


### Establishing a Connection

```{r}
con <- DBI::dbConnect(RSQLite::SQLite(), dbname = "nba.sqlite")
```

# Exploratory Data Analysis

## Question 1: How many players went UCLA

```{r}
dbGetQuery(con, "SELECT COUNT(*) AS ucla_alumni
                 FROM common_player_info
                 WHERE school LIKE 'UCLA'
           ")
```

\newpage

## Question 2: What is the Average Draft Combine Statistics Overtime

```{r}
combine_stats_avg <- dbGetQuery(con, "SELECT season, 
                                             AVG(height_wo_shoes) AS avg_height, 
                                             AVG(weight) AS avg_weight,
                                             AVG(wingspan) AS avg_wingspan,
                                             AVG(standing_reach) AS avg_standing_reach,
                                             AVG(standing_vertical_leap) 
                                             AS avg_standing_vertical_leap,
                                             AVG(max_vertical_leap) AS avg_max_vertical_leap, 
                                             AVG(lane_agility_time) AS avg_lane_agility_time,
                                             AVG(modified_lane_agility_time) 
                                             AS modified_lane_agility_time,
                                             AVG(three_quarter_sprint) AS avg_three_quarter_sprint,
                                             AVG(bench_press) AS avg_bench_press
                                      FROM draft_combine_stats
                                      GROUP BY season
                                       
                                 ")

head(combine_stats_avg)
combine_stats_avg$season <- as.integer(combine_stats_avg$season)

```

```{r}
# Pivot to long format
combine_stats_long <- combine_stats_avg %>%
  pivot_longer(
    cols = -season,
    names_to = "metric",
    values_to = "average"
  )

combine_stats_long$season <- as.integer(combine_stats_long$season)


# Plot all metrics in one line chart with color and legend
ggplot(combine_stats_long, aes(x = season, y = average, color = metric)) +
  geom_line(size = .5) +
  # geom_point(size = .25) +
  theme_minimal() +
  labs(
    title = "Average NBA Combine Metrics by Season",
    x = "Season",
    y = "Average Value",
    color = "Metric"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Analysis Part 1

Here we can see all the average values for the combine statistics of all the players from 2000 up until now. Although each line tends to be relatively straight, let's anayze these metrics in their own plots.

```{r}
generate_plots <- function(data) {
  column_names <- colnames(data)[-1]
  
  data$season <- as.integer(data$season)
  plot_list <- list()

  # Code to execute
  for (col in column_names) {
     plot <- ggplot(data, aes(x = season, y = .data[[col]])) +
            geom_smooth(color = "lightblue", size = .5) +
            theme_minimal(base_size = 4.5) +
            labs(
              title = paste(col, "at NBA Draft Combine by Season"),
              x = "Season",
              y = col
            ) +
            theme(axis.text.x = element_text(angle = 45, hjust = 1))
     
    plot_list[[col]] <- plot
  }
  return(plot_list)
}

# Generate the plots
plot_list <- generate_plots(combine_stats_avg)
Reduce(`+`, plot_list) + plot_layout(ncol = 2)
```

#### Analysis Part 2

When looking at the individual plots for the metrics, some metrics have a apparent change over time. These most noticable include...

- The average weight of players has went down 
- The average max vertical leap of players went up
- The average lane agility time of players went down
- The average three quarter sprint time of players went down
- The average amount of benchpress reps of players went down

This shows how each draft class of NBA players are becoming more athletic. Players are jumping higher, running faster, and becoming more leaner.

\newpage

## Question 3: What is the Average Draft Combine Statistics for the Top 10 Overall Picks Each Year

```{r}
combine_stats_avg_top_10 <- dbGetQuery(con, "WITH draftStatsMetrics AS (
                                        SELECT dcs.season, dcs.player_id, dcs.player_name, 
                                               dh.overall_pick, dcs.position, 
                                               dcs.height_wo_shoes, dcs.weight, 
                                               dcs.wingspan, dcs.standing_reach,
                                               dcs.standing_vertical_leap, 
                                               dcs.max_vertical_leap, 
                                               dcs.lane_agility_time,
                                               dcs.three_quarter_sprint
                                        FROM draft_combine_stats AS dcs
                                        INNER JOIN draft_history AS dh
                                        ON dcs.player_id = dh.person_id
                                     )
                    
                                  SELECT season, 
                                         AVG(height_wo_shoes) AS avg_height, 
                                         AVG(weight) AS avg_weight,
                                         AVG(wingspan) AS avg_wingspan,
                                         AVG(standing_reach) AS avg_standing_reach,
                                         AVG(standing_vertical_leap) 
                                          AS avg_standing_vertical_leap,
                                         AVG(max_vertical_leap) 
                                          AS avg_max_vertical_leap, 
                                         AVG(lane_agility_time) 
                                          AS avg_lane_agility_time,
                                         AVG(three_quarter_sprint) 
                                          AS avg_three_quarter_sprint
                                  FROM draftStatsMetrics
                                  WHERE overall_pick <= 10
                                  GROUP BY season
                                  ")

head(combine_stats_avg_top_10)
```

```{r}
# Pivot to long format
combine_stats_long_top_10 <- combine_stats_avg_top_10 %>%
  pivot_longer(
    cols = -season,
    names_to = "metric",
    values_to = "average"
  )

combine_stats_long_top_10$season <- as.integer(combine_stats_long_top_10$season)

# Plot all metrics in one line chart with color and legend
ggplot(combine_stats_long_top_10, aes(x = season, y = average, color = metric)) +
  geom_line(size = .5) +
  # geom_point(size = .25) +
  theme_minimal() +
  labs(
    title = "Average NBA Combine Metrics by Season",
    x = "Season",
    y = "Average Value",
    color = "Metric"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Analysis Part 1

Again, here is a plot of all the average combine metrics for the Top 10 Players overtime. These lines are not indicative to how sensitive these metrics are. Let's analyze the individual metric plots again.

```{r}
plot_list <- generate_plots(combine_stats_avg_top_10)
Reduce(`+`, plot_list) + plot_layout(ncol = 2)
```

#### Analysis Part 2

From these plots we can see that the Top 10 Players in the draft class has become athletic in terms of...

- Higher Average Standing Vertical Leap
- Higher Average Max Vertical Leap
- Lower Average Lane Agility Time

\newpage

## Question 4: Do players of certain positions commit more fouls?

```{r}
position_fouls <- dbGetQuery(con, "WITH fouls AS (
                                     SELECT game_id, eventnum, eventmsgtype, 
                                            homedescription, player1_id, player1_name, 
                                            player1_team_id
                                     FROM play_by_play
                                     WHERE homedescription LIKE '%Foul%'
                                     UNION ALL 
                                     SELECT game_id, eventnum, eventmsgtype, 
                                            visitordescription, player2_id, player2_name, 
                                            player2_team_id
                                     FROM play_by_play
                                     WHERE visitordescription LIKE '%Foul%'
                                  ),
                                  
                                  position AS (
                                     SELECT person_id, display_first_last, position
                                     FROM common_player_info
                                  ),
                                  
                                  season AS (
                                     SELECT SUBSTRING(season_id, 2) AS current_season, 
                                     game_id, season_type
                                     FROM game
                                  ),
                                  
                                  foulCounts AS (
                                    SELECT current_season, position, season_type, 
                                    COUNT(*) AS fouls
                                    FROM fouls AS f
                                    INNER JOIN position AS p
                                    ON f.player1_id = p.person_id
                                    INNER JOIN season AS s
                                    ON f.game_id = s.game_id
                                    WHERE season_type = 'Regular Season' 
                                      AND position != ''
                                    GROUP BY current_season, position
                                  ),
                                  
                                  positionGroup AS (
                                    SELECT *
                                    FROM foulCounts
                                  ),
                                  
                                  foulSum AS (
                                    SELECT current_season, SUM(fouls) AS fouls, position
                                    FROM positionGroup
                                    GROUP BY current_season, position
                                    ORDER BY current_season, position
                                  ),
                                  
                                  foulTotal AS (
                                    SELECT *,
                                           SUM(fouls) OVER (
                                              PARTITION BY current_season
                                           ) AS fouls_total
                                    FROM foulSum
                                  )
                                  
                                  SELECT *,
                                         ((1.0 * fouls) / fouls_total) * 100 AS position_pct
                                  FROM foulTotal
           ")

head(position_fouls)
```

#### Data Cleaning

```{r}
position_fouls <- position_fouls %>%
  mutate(current_season = as.integer(current_season))
```

#### Visualization

```{r}
ggplot(position_fouls, aes(x = current_season, y = position_pct, color = position)) +
  geom_line(size = .5) +
  # geom_point(size = .25) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    title = "Percentage of Fouls by Position Over Seasons",
    x = "Season",
    y = "Percentage of Fouls",
    color = "Position"
  ) +
  theme_minimal()
```

#### Analysis Part 1

Here we can see how each position in the NBA contributes to the total percentage of fouls per year overtime. We can see that although players at the Guard position didn't account for the highest percentage of fouls initially, we can see that over time their contribution to the number of fouls increases drastically. We can see the opposite relationship when analyzing players playing at the center position. They first account for ~18% of fouls initially but has starkly decreased to around ~12% in recent years. We can see these trends better if we generalize these positions to the positions that these players favor (Center, Guard, Forward).

### Generalize to Centers, Guards, and Forwards Only

```{r}
position_fouls <- dbGetQuery(con, "WITH fouls AS (
                                     SELECT game_id, eventnum, eventmsgtype, homedescription, 
                                            player1_id, player1_name, player1_team_id
                                     FROM play_by_play
                                     WHERE homedescription LIKE '%Foul%'
                                     UNION ALL 
                                     SELECT game_id, eventnum, eventmsgtype, visitordescription, 
                                            player2_id, player2_name, player2_team_id
                                     FROM play_by_play
                                     WHERE visitordescription LIKE '%Foul%'
                                  ),
                                  
                                  position AS (
                                     SELECT person_id, display_first_last, position
                                     FROM common_player_info
                                  ),
                                  
                                  season AS (
                                     SELECT SUBSTRING(season_id, 2) AS current_season, 
                                     game_id, season_type
                                     FROM game
                                  ),
                                  
                                  foulCounts AS (
                                    SELECT current_season, position, season_type, 
                                    COUNT(*) AS fouls
                                    FROM fouls AS f
                                    INNER JOIN position AS p
                                    ON f.player1_id = p.person_id
                                    INNER JOIN season AS s
                                    ON f.game_id = s.game_id
                                    WHERE season_type = 'Regular Season' AND position != ''
                                    GROUP BY current_season, position
                                  ),
                                  
                                  positionGroup AS (
                                    SELECT *,
                                          CASE
                                             WHEN position LIKE 'Guard%' THEN 'Guard'
                                             WHEN position LIKE 'Forward%' THEN 'Forward'
                                             WHEN position LIKE 'Center%' THEN 'Center'
                                          END AS position_group
                                    FROM foulCounts
                                  ),
                                  
                                  foulSum AS (
                                    SELECT current_season, SUM(fouls) AS fouls, 
                                    position_group
                                    FROM positionGroup
                                    GROUP BY current_season, position_group
                                    ORDER BY current_season, position_group
                                  ),
                                  
                                  foulTotal AS (
                                    SELECT *,
                                           SUM(fouls) OVER (
                                              PARTITION BY current_season
                                           ) AS fouls_total
                                    FROM foulSum
                                  )
                                  
                                  SELECT *,
                                         ((1.0 * fouls) / fouls_total) * 100 AS position_pct
                                  FROM foulTotal
           ")

head(position_fouls)
```

#### Data Cleaning

```{r}
position_fouls <- position_fouls %>%
  mutate(current_season = as.integer(current_season))
```

#### Visualization

```{r}
ggplot(position_fouls, aes(x = current_season, y = position_pct, color = position_group)) +
  geom_line(size = .5) +
  # geom_point(size = .25) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    title = "Percentage of Fouls by Position Over Seasons",
    x = "Season",
    y = "Percentage of Fouls",
    color = "Position"
  ) +
  theme_minimal()
```

#### Analysis Part 2

In this graph we generalized the different types of positions to the main 3 (Center, Forward, and Guard). Here we can clearly see players playing at the Center position have contributed much less to the number of fouls while guards grown to account for the most interchangeable with the forward positon in recent years.                                

\newpage

## Question 5: Do certain schools have a higher chance of producing players of a certain position?

#### SQL Query

```{r}
school_position <- dbGetQuery(con, "WITH playerPosition AS (
                                      SELECT *,
                                             CASE
                                               WHEN position LIKE '%Guard%' THEN 'Guard'
                                               WHEN position LIKE '%Forward%' AND position 
                                                NOT LIKE '%Guard%' THEN 'Forward'
                                               WHEN position LIKE '%Center%' AND position 
                                                NOT LIKE '%Forward%' THEN 'Center'
                                               ELSE 'Other'
                                             END AS position_group
                                      FROM common_player_info AS cpi
                                      LEFT JOIN draft_history AS dh
                                      ON cpi.person_id = dh.person_id
                                      ),
                                   
                                   positionCount AS (
                                      SELECT school, position_group, 
                                             COUNT(*) AS n_position,
                                             SUM(COUNT(*)) OVER (
                                             PARTITION BY school
                                             ) AS school_total_draft
                                      FROM playerPosition
                                      WHERE school NOT NULL AND school != '' 
                                        AND school NOT LIKE '% %' AND position != ''
                                      GROUP BY school, position_group
                                      ORDER BY n_position, position_group
                                   )
                                   
                                   SELECT school, position_group AS position, 
                                          n_position, school_total_draft, 
                                          ROUND(1.0 * n_position / school_total_draft, 3) 
                                            AS pct_school
                                   FROM positionCount
                                   ORDER BY school_total_draft DESC
                                   
                                   
                             ")
head(school_position)
```

#### Top 10 Schools

```{r}
top_schools <- school_position %>%
  group_by(school) %>%
  summarise(total = sum(n_position)) %>%
  top_n(10, total) %>%
  pull(school)
```

#### Visualization: Stacked Bar Chart

```{r}
ggplot(
  school_position %>% filter(school %in% top_schools),
  aes(x = reorder(school, -school_total_draft), y = pct_school, fill = position)
) +
  geom_bar(stat = "identity") +
  labs(
    title = "Proportion of Player Positions by School",
    x = "School", y = "Proportion of Players",
    fill = "Position"
  ) +
  scale_fill_manual(values = c("Guard" = "#17408B", "Forward" = "#c8102e", "Center" = "#FAC54E")) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

#### Analysis Part 1

Here we can see the Top 10 schools where players come from when entering the NBA draft. We can see that schools like Arizona, Indiana, Villanova, Louisville, Kentucky output more than 50% of their athletes as Guards to the draft. 


#### Visualization: Grouped Bar Charts

```{r}
ggplot(
  school_position %>% filter(school %in% top_schools),
  aes(x = reorder(school, -school_total_draft), y = pct_school, fill = position)
) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Position Breakdown by School",
    x = "School", y = "Proportion",
    fill = "Position"
  ) +
  scale_fill_manual(values = c("Guard" = "#17408B", "Forward" = "#c8102e", "Center" = "#FAC54E")) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

#### Analysis Part 2

Here we can see better how the Top 10 schools compare to their players' positions when entering the draft. Arizona outputs the most Guards, Syracuse outputs the most Centers, adn Michigan outputs the mose Forwards.

\newpage

## Question 6: Is there a relationship between the characteristics of a player and the position(s) that they play?

#### SQL Query

```{r}
position_combine <- dbGetQuery(con, "WITH playerPosition AS (
                                      SELECT cpi.person_id, 
                                             cpi.display_first_last, dcs.position,
                                             CASE
                                               WHEN dcs.position LIKE 'PG%' 
                                                THEN 'Point Guard'
                                               WHEN dcs.position LIKE 'PF%' 
                                                THEN 'Power Forward'
                                               WHEN dcs.position LIKE 'SG%' 
                                                THEN 'Shooting Guard'
                                               When dcs.position LIKE 'SF%' 
                                                THEN 'Small Forward'
                                               WHEN dcs.position LIKE 'C%' 
                                                THEN 'Center'
                                               ELSE 'Other'
                                             END AS position_group,
                                            dcs.height_wo_shoes, dcs.weight, 
                                            dcs.wingspan, dcs.standing_reach,
                                            dcs.standing_vertical_leap, dcs.max_vertical_leap, 
                                            dcs.lane_agility_time,
                                            dcs.three_quarter_sprint
                                      FROM common_player_info AS cpi
                                      RIGHT JOIN draft_combine_stats AS dcs
                                      ON cpi.person_id = dcs.player_id
                                      WHERE dcs.position NOT NULL AND dcs.position !=''
                                      )
                         
                                SELECT *
                                FROM playerPosition
                                
                             ")

head(position_combine)
```

#### Visualization: Box-Whisker Plots

```{r}
position_combine_cleaned <- position_combine %>%
  drop_na(height_wo_shoes, weight, wingspan, standing_reach, standing_vertical_leap,
          max_vertical_leap, lane_agility_time, three_quarter_sprint) %>%
    mutate(across(c(height_wo_shoes, weight, wingspan, standing_reach, 
                  standing_vertical_leap, max_vertical_leap, 
                  lane_agility_time, three_quarter_sprint), 
                ~ as.numeric(.)))

nrow(position_combine_cleaned)
```

```{r}
make_boxplots <- function(data, position_col, variables) {
  
  plot_list <- list()
  
  for (var in variables) {
    p <- ggplot(data, aes_string(x = position_col, y = var, fill = position_col)) +
      geom_boxplot(outlier.shape = 10) +
      labs(title = var,
           x = "Position",
           y = var) +
      theme_minimal(base_size = 8) +  # slightly smaller text for better layout
      theme(axis.text.x = element_text(angle = 90, hjust = 1),
            legend.position = "none")
    
    plot_list[[var]] <- p
  }
  
  # Combine plots using patchwork
  combined_plot <- wrap_plots(plot_list, ncol = 4) 
  print(combined_plot)
}

vars_to_plot <- c("height_wo_shoes", "weight", "wingspan", 
                  "standing_reach", "standing_vertical_leap",
                  "max_vertical_leap", "lane_agility_time", "three_quarter_sprint")

make_boxplots(position_combine_cleaned, position_col = "position_group", variables = vars_to_plot)
```

#### Analysis

Here we can see the metrics that divide the different playing positions. Select positions have many metrics that indicate their ideal position. Centers tend to be the tallest, heaviest, possess the largest wingspan, tallest standing reach, lowest standing vertical leap, lowest max vertical leap, slowest lane agility time, and the slowest three quarter sprint. Guards on the other hand tend to be the shortest, weight the least, have the smallest wingspan, have the smallest standing_reach, have the highest max vertical leap, have the fastest lane agility time, and the fastest three quarter sprint.

This graphs support the skill set needed by each playing position. Centers typically are expected to play more within the paint, grabbing rebounds by utilizing their big frame to their advantage. Point Guards use their smaller frame to speed past opponents and look for opportunities to execute plays for themselves and their team.

\newpage

## Question 7: Which Team has the Best Regualar Season Winning Percentage

#### SQL Query

```{r}
best_teams <- dbGetQuery(con, "WITH combinedTeam AS (
                     SELECT SUBSTRING(season_id, 2) AS season, 
                     team_name_home AS team_name, wl_home,
                        CASE 
                          WHEN wl_home = 'W' THEN 1
                          ELSE 0
                        END AS win_count,
                        CASE 
                          WHEN wl_home = 'L' THEN 1
                          ELSE 0
                        END AS loss_count,
                        season_type
                     FROM game
                     UNION ALL
                     SELECT SUBSTRING(season_id, 2) AS season, 
                     team_name_away AS team_name, wl_away,
                        CASE 
                          WHEN wl_away = 'W' THEN 1
                          ELSE 0
                        END AS win_count,
                        CASE 
                          WHEN wl_away = 'L' THEN 1
                          ELSE 0
                        END AS loss_count,
                        season_type
                     FROM game
                ),
                
                teamRecords AS (
                    SELECT season, team_name, SUM(win_count) AS season_wins, 
                    SUM(loss_count) AS season_losses
                    FROM combinedTeam
                    WHERE season_type = 'Regular Season'
                    GROUP BY season, team_name
                )
                
                SELECT *, (1.0 * season_wins ) / (season_wins + season_losses) AS win_pct
                FROM teamRecords
                ORDER BY win_pct DESC
                LIMIT 10
           ")

best_teams
```

#### Visualization: Bar Chart

```{r}

# Order team names by win percentage
best_teams$team_label <- paste(best_teams$team_name, best_teams$season, sep = ": ")
best_teams$record <- paste0(best_teams$season_wins, "-", best_teams$season_losses)

team_colors <- c(
  "Golden State Warriors" = "#FDB927",
  "Chicago Bulls" = "#CE1141",
  "Los Angeles Lakers" = "#552583",
  "Boston Celtics" = "#007A33",
  "Dallas Mavericks" = "#00538C"
)



# Plot
ggplot(best_teams, aes(x = reorder(team_label, win_pct), y = win_pct, fill = team_name)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = record), 
            hjust = -0.1, 
            color = "black", 
            size = 3.5) +
  coord_flip() +
  scale_fill_manual(values = team_colors) +
  labs(title = "Top NBA Teams by Season Win Percentage",
       x = "Team (Season)",
       y = "Win Percentage") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1), expand = expansion(mult = c(0, 0.1))) +
  theme_minimal() +
  theme(legend.position = "none")

```

#### Analysis

Here we can see the top 10 teams with the best regular season winning percentage of all time. The best team of all time is the 2015 Golden State Warriors.

\newpage

## Question 8: Which Teams has the Best Regualar Season Winning Percentage Over a Four Year Period?

```{r}
best_team_4_years <- dbGetQuery(con, "WITH combinedTeam AS (
                     SELECT SUBSTRING(season_id, 2) AS season, 
                     team_name_home AS team_name, wl_home,
                        CASE 
                          WHEN wl_home = 'W' THEN 1
                          ELSE 0
                        END AS win_count,
                        CASE 
                          WHEN wl_home = 'L' THEN 1
                          ELSE 0
                        END AS loss_count,
                        season_type
                     FROM game
                     UNION ALL
                     SELECT SUBSTRING(season_id, 2) AS season, 
                     team_name_away AS team_name, wl_away,
                        CASE 
                          WHEN wl_away = 'W' THEN 1
                          ELSE 0
                        END AS win_count,
                        CASE 
                          WHEN wl_away = 'L' THEN 1
                          ELSE 0
                        END AS loss_count,
                        season_type
                     FROM game
                ),
                
                teamRecords AS (
                    SELECT season, team_name, SUM(win_count) AS season_wins, 
                    SUM(loss_count) AS season_losses
                    FROM combinedTeam
                    WHERE season_type = 'Regular Season'
                    GROUP BY season, team_name
                ),
                
                fourYearPeriod AS(
                    SELECT *, (1.0 * season_wins ) / (season_wins + season_losses) AS win_pct,
                           SUM(season_wins) OVER (
                              PARTITION BY team_name
                              ORDER BY season ASC, team_name ASC
                              ROWS BETWEEN 3 PRECEDING AND CURRENT ROW
                           ) AS season_wins_4_year,
                           SUM(season_losses) OVER (
                              PARTITION BY team_name
                              ORDER BY season ASC, team_name ASC
                              ROWS BETWEEN 3 PRECEDING AND CURRENT ROW
                           ) AS season_losses_4_year,
                          COUNT(*) OVER (
                              PARTITION BY team_name
                              ORDER BY season ASC, team_name ASC
                              ROWS BETWEEN 3 PRECEDING AND CURRENT ROW
                           ) AS count_years
                    FROM teamRecords
                )
                
                SELECT season, team_name, season_wins, season_losses, 
                       win_pct, season_wins_4_year, season_losses_4_year,
                       (1.0 * season_wins_4_year) / (season_wins_4_year + season_losses_4_year) 
                          AS win_pct_4_year,
                       count_years
                FROM fourYearPeriod
                WHERE count_years = 4
                ORDER BY win_pct_4_year DESC
                LIMIT 10
                
           ")

best_team_4_years
```

#### Transform Data

```{r}
# First, ensure season column is integer
best_team_4_years <- best_team_4_years %>%
  mutate(season = as.integer(season))

# Now transform season into "start-end" format
best_team_4_years <- best_team_4_years %>%
  mutate(season_window = paste0(season - 4, "-", season))

best_team_4_years
```

```{r}
# Order team names by win percentage
best_team_4_years$team_label <- paste(best_team_4_years$team_name, best_team_4_years$season_window, sep = ": ")
best_team_4_years$record <- paste0(best_team_4_years$season_wins_4_year, "-", best_team_4_years$season_losses_4_year)

# Plot
ggplot(best_team_4_years, aes(x = reorder(team_label, win_pct), y = win_pct, fill = team_name)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = record), 
            hjust = -0.1, 
            color = "black", 
            size = 3.5) +
  coord_flip() +
  scale_fill_manual(values = team_colors) +
  labs(title = "Top NBA Teams by Four Year Season Win Percentage",
       x = "Team (Season Period)",
       y = "Win Percentage") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1), expand = expansion(mult = c(0, 0.1))) +
  theme_minimal() +
  theme(legend.position = "none")
```

#### Analysis

Here we can see the top 10 teams by a four year period winning percentage of all time. The Golden State Warriors claim the highest winning percentage with a 258-70 record.

\newpage

## Question 9: What Player has the Most Fouls of All Time

```{r}
playerFouls <- dbGetQuery(con, "WITH fouls AS (
                                   SELECT game_id, eventnum, eventmsgtype, 
                                          homedescription, player1_id, player1_name, 
                                          player1_team_id
                                   FROM play_by_play
                                   WHERE homedescription LIKE '%Foul%'
                                   UNION ALL 
                                   SELECT game_id, eventnum, eventmsgtype, 
                                          visitordescription, player2_id, player2_name, 
                                          player2_team_id
                                   FROM play_by_play
                                   WHERE visitordescription LIKE '%Foul%'
                              )
                              
                              SELECT player1_id AS player_id, player1_name AS player_name, \
                                     COUNT(*) AS foul_count
                              FROM fouls
                              WHERE player_id != '0'
                              GROUP BY player1_id, player1_name
                              ORDER BY COUNT(*) DESC
                              LIMIT 10
           ")

playerFouls
```

```{r}
# Create alternating color vector
alternating_colors <- rep(c("#17408B", "#c8102e"), length.out = nrow(playerFouls))


# Create the bar chart
ggplot(playerFouls, aes(x = player_name, y = foul_count, fill = player_name)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = foul_count), vjust = -0.5, size = 2.5) +
  scale_fill_manual(values = alternating_colors) +
  labs(title = "Top 10 NBA Players by Total Fouls",
       x = "Player",
       y = "Total Fouls") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

#### Analysis

Here we can see the player that has committed the most fouls of all time is Dwight Howard.







